---
output: word_document  
---

This document contains settings that help exporting your analysis to a Word document. The expectation is that you have already gained familiarity with editing R Markdown documents from the [R for Data Science](https://r4ds.had.co.nz/r-markdown.html) book, which was on the module reading list, particularly Chapters 27, 29 and 30. It assumes, therefore, that you are familiar with the purpose of yaml headers (i.e. the settings between the two sets of - - - marks at the top of an `.Rmd` document). There are numerous settings that can be included in the yaml header, but for our purposes specifying the output format (Word document) is the only essential one.

Options relating to how code chunks (i.e. the commands included between two sets of \`\`\` marks throughout) are rendered when written (i.e. "knitted") to the output document format can be set both in the header of the individual code chunks, or as a global setting that applies to all chunks. The first code chunk below (in the `.Rmd` version) is a global setting chunk that applies some useful setting for the purpose of printing outputs only to Word documents.

To learn more about R Markdown documents, these two resources are extremely useful: 

  * Xie, Allaire, and Grolemund (2021) [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/) 
  
  * Xie, Dervieux, and Riederer (2022) [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)

These resources go far beyond what you need to know at this stage, but they can be helpful for searching particular settings/issues that come up in your own coding.

Below is a recommended structure and some example code for tabulations that seems to work well with Word documents.

```{r Setup, include=FALSE}
# Keep these options as they appear to produce the best results when kniting to Word documents

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      results = 'asis', render = knitr::normal_print)
```

```{r include=FALSE}

# Load/install necessary packages using the 'pacman' package;
# install.packages("pacman") # first, if 'pacman' is not yet installed

if (!require("pacman")) install.packages("pacman") ## installs 'pacman' if not yet installed

pacman::p_load(
  tidyverse,    # general data management tools ('dplyr', etc.)
  mosaic,       # formula-type syntax for descriptive statistics
  ggformula,    # ggplot2 powered graphing using 'mosaic' formula-syntax
  lme4,         # multilevel modelling
  summarytools, # summary statistics tables
  sjlabelled,   # data import from other software (alternative to 'haven') and labels management
  sjmisc,       # data transformation on variables (recoding,grouping, missing values, etc)
  sjPlot,       # Graphing and tabulation tools for regression model results
  jtools,       # Graphing and tabulation tools for regression model results
  parameters    # ibid
  )
```

# Your title

## Introduction

## Literature review (or a better title for this section)

### Any subtitles

## Data and methods

```{r import data, include=F}
# Loads and prepares example data for the tables

ost <- sjlabelled::read_stata("https://cgmoreh.github.io/SSC7001M/data/osterman.dta") %>%
          select("trustindex3", "ppltrst", "pplfair", "cntry", "facntr", "mocntr", "female", "agea", "eduyrs25", "paredu_a_high") %>%
          filter(cntry %in% c("GB", "IE", "DE", "FR", "HU", "PL", "PT", "ES")) %>%  
          mutate(across(c("ppltrst", "pplfair"), sjlabelled::as_label),
                cntry = as_factor(cntry),
                 fmnoncntr = ifelse(facntr==0 | mocntr==0, 1, 0)) %>%
          sjlabelled::var_labels(trustindex3 = "Social trust scale",
                                 eduyrs25 = "Years of full-time education",
                                 paredu_a_high = "High parental education",
                                 fmnoncntr = "Least one parent born abroad")
```


Somewhere in this section or early on in the Results section you may want to show a summary of your variables of interest. For example, a table such as this (using the `descr()` function from the `summarytools` package):


```{r}
## 1. Take a dataset %>%  
##    2. Treat variables as numeric for the purpose of the table %>% 
##    3. Optionally, select the variables to include in the table; add variable names inside the (), using quotation marks and separated by commas %>% 
##    4. run the descr() command to create the table

ost %>% 
  as_numeric() %>%
  select("trustindex3", "ppltrst", "pplfair", "cntry", "facntr", "mocntr", "female", "agea", "eduyrs25", "paredu_a_high") %>%  
  summarytools::descr(stats = c("mean", "sd", "med", "min", "max", "n.valid"),
      transpose = TRUE,
      silent = TRUE,
      headings  = FALSE,
      )
```

The main aesthetic limitation is that the number of decimal points cannot be set separately for each statistic, which in this particular table is inconvenient. Decimals can be rounded down to 0, but then the Mean and Std. Dev also gets rounded down.Also, variables are re-ordered alphabetically.

Another option is using another function also called `descr()`, but from the `sjmisc` package:

```{r}
# Note that the results of the 'descr()' command need to be passed on to 'print_md()' for it to look good in Word
# It may also be useful to first treat some variables as of certain types (e.g. as_factor(), as_numeric() as below)

ost %>% as_factor() %>%  as_numeric(trustindex3, agea, eduyrs25) %>% 
  select("trustindex3", "ppltrst", "pplfair", "cntry", "facntr", "mocntr", "female", "agea", "eduyrs25", "paredu_a_high") %>%  
  sjmisc::descr(show = c("n", "NA.prc", "mean", "sd", "md", "range")) %>% print_md()
```

Here, on the other hand, what's inconvenient is that the order of the statistics (column headings) cannot be changed. That's probably easier to do manually in Word, however, if needed.

## Results

This is the section where you describe the results of your main analysis, which will be some regression model. In this example, I fit three nested models - M1, M2 and M3 - starting with a simple bivariate regression of `trust` on `education years` only (M1), then expanding the list of predictors with `age` and `sex` (M2), and finally including also `parental education` and `parents born abroad`. The *`.Rmd`* file fits and saves these models in the background. 

In class, your aim was to print summary statistics from the fit model in a format that helped you to interpret the results. Apart from the standard call to `summary()`, we have tried `summ()` from the `jtools` package.

Here, your aim is to find the best way to present your model(s) to readers. The two summary tools we used in class are not ideal for directly knitting nicely formatted regression model summaries to Word documents in RMarkdown. The *`.Rmd`* file contains some better functions for this purpose.

```{r echo=FALSE}
# A linear model with a single predictor

M1 <- lm(trustindex3 ~ eduyrs25, data = ost)
```

```{r echo=FALSE}
# A multiple regression model with some demographic predictors
M2 <- lm(trustindex3 ~ eduyrs25 + agea + female, data = ost)
```

```{r echo=FALSE}
# A multiple regression model with several other important predictors
M3 <- lm(trustindex3 ~ eduyrs25 + agea + female + paredu_a_high + fmnoncntr, data = ost)
```

Here is an example of a summary table of a single model using the `parameters` package):

```{r}
## The only thing to change is the model name (M3)
parameters::model_parameters(M3, digits = 3) %>% 
  print_md()
```


A summary table of several models (using the `parameters` package):

```{r}
parameters::compare_models(M1, M2, M3, style = "se_p") %>%
  print_md()
```

## Discussion

## Conclusions

## References
