---
title: "Week 3: Simple linear regression"
output:
  rmdformats::readthedown:
      highlight: tango
---
```{r include=FALSE}
knitr::opts_chunk$set(
  fig.width = 10, fig.height = 6, fig.retina = 2,
  warning = FALSE, message = FALSE, eval=T)
```



# Introduction
- What are the aims of regression?
  - Regression for prediction
  - Regression for causal analysis
  
# Prediction

Let's look at a dataset of Boris Johnson's public opinion approval before and after the COVID lockdowns. 
We have a set of predictor variables: (1) COVID deaths over time and (2) the monthly unemployment rate.

```{r}
bj <- readRDS(url("https://cgmoreh.github.io/SSC7001M/data/BorisJohnson_Approval2.rds"))
```

## Fit a normal linear regression model

mod <- lm(Net_Positive ~ Unemployment_Rate_Rounded + Monthly_Deaths_10k, data = bj)
summary(mod)



| Name            | Description                                                                                                  |
|-----------------|--------------------------------------------------------------------------------------------------------------|
| `to_labour`       | Did respondent read a newspaper that switched endorsement to the Labour between 1992 and 1997 (1=Yes, 0=No)? |
| `vote_lab_92`   | Did respondent vote for Labour in 1992 election (1=Yes, 0=No)?                                               |
| `vote_lab_97`   | Did respondent vote for Labour in 1997 election (1=Yes, 0=No)?                                               |
| `parent_labour` | Did the respondent's parents vote for Labour (1=Yes, 0=No)?                                                  |
| `wkclass`       | Does the respondent identify as working class (1=Yes, 0=No)?                                                 |
| `male`          | Does the respondent identify as Male (1=Yes, 0=No)?                                                          |
| `age`     | Age of respondent |


The treatment here is someone reading a paper that switched their endorsement. The outcome of interest is how the respondent voted in 1997.


```{r setup}
news <- read.csv("data/newspapers.csv")
```
