---
title: "Week 6: Multicategory regression models"
output: 
  rmdformats::readthedown:
    code_folding: hide
    code_download: true
    highlight: tango
    number_sections: false
---
```{r include=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```


# Introduction

This session introduces two new regression models:
- Ordinal logistic regression, and
- Multinomial logistic regression
  
# Ordinal Regression 

Ordinal regression (also known as Ordinal Logistic Regression) is an extension of binomial logistics regression. Ordinal regression is used to predict a dependent variable with ‘ordered’ multiple categories and independent variables. For example: let us assume that in a survey we asked a question with answer options such as Strongly Disagree, Disagree, Agree, Strongly Agree. This helped us to observe a natural order in the categories and for our regression model to be realistic, we must appreciate this order. Ordinal Logistic Regression addresses this fact. Ordinal means that there is a logical order to the categories.

One rule of thumb that is popular in the social and behavioral sciences is to treat an ordinal variable as continuous if it has seven or more categories (some claim five is sufficient). However, this approach depends on the distribution of the categories. If the distribution does not approximate a normal distribution, then biased estimates are likely. Moreover, we must make a strong assumption about the “distance” between categories. When measuring, say, age in years, we know that the distance between 10 and 20 is 10 times the distance between 1 and 2. For ordinal variables, this is rarely, if ever, the case (what is the distance between agree and disagree?). Given these constraints, most analysts prefer to use a regression model that is designed specifically for ordinal outcomes.


# Multinomial regression

Multinomial Logistic Regression (MLR) is a form of linear regression analysis conducted when the dependent variable is nominal with more than two levels. It is used to describe data and to explain the relationship between one dependent nominal variable and one or more continuous-level (interval or ratio scale) independent variables. You can understand nominal variable as, a variable which has no intrinsic ordering.

For example: Types of Forests: ‘Evergreen Forest’, ‘Deciduous Forest’, ‘Rain Forest’. As you see, there is no intrinsic order in them, but each forest represent a unique category. In other words, multinomial regression is an extension of logistic regression, which analyzes dichotomous (binary) dependents.

# Fitting multinomial and ordinal regression modesl in R

There are several packages in `R` that provide functions for fitting multicategory models, such as the `nnet` and `mlogit` packages for multinomial models, the `MASS` package for ordinal models, or `vgl`, which fits both. Each of these packages have their advantages and disadvantages in terms of the format of their outputs, summary statistic provided, etc. 

# Load useful packages

```{r}
pacman::p_load(tidyverse, VGAM, ordinal, nnet, stargazer, magrittr, brant, MASS)
```



# Data import and wrangling

```{r}
# If we have already checked a variable list or questionnaire and we know which variables we want to import from a larger dataset, we can select only those variables using the 'col_select' option rom the 'haven' package

evs <- haven::read_dta("https://cgmoreh.github.io/SSC7001M/data/evs5.dta", col_select = c("v31":"v37", "v225", "v234", "v227", "age", "v102", "v105"))
```

```{r}
# We can create a data dictionary to quickly check variable names and labels

evs_dictionary <- labelled::generate_dictionary(evs) 
```

```{r eval=FALSE, include=FALSE}
# We can select some variables of interest that we would like to model even if we had previously imported a larger dataset

# evs <- evs %>% select("v31":"v37", "v225", "v234", "v227", "age", "v102", "v105")
```

```{r eval=FALSE, include=FALSE}
# We can transform the imported categorical data from 'haven labelled' format to 'factors' (categorical variables)

evs <- evs %>% mutate_if(haven::is.labelled, haven::as_factor)
```


```{r}
# We can check the frequencies of some variables; 
## replace the variable name to check others;
## come back and check again after recoding the variables later

evs %>% count(v227)

```


```{r}
# We can set some of the non-relevant categories to be missing; in this case, this applies to all the variables;
## for simplicity, we use the %<>% pipe from the magrittr package, which saves us some repetitive typing

evs %<>% mutate(across(.cols = everything(), na_if, "multiple answers Mail"))
evs %<>% mutate(across(.cols = everything(), na_if, "no answer"))
evs %<>% mutate(across(.cols = everything(), na_if, "dont know"))
```



```{r}
# evs <- evs %>% 
#  mutate(across(c("v32":"v37"), as_factor),
#  mutate(across(c("v31", "v225", "v234", "v227"), as_factor)),
#  mutate(across(c("age", "v102", "v105"), as.numeric)))
```


## Binomial logistic regression

We can first fit a binary logistic regression model to predict general social trust (`v31`) from the following predictor variables:
- `v225`: gender
- `age`: age
- `v227`: whether born in the country of data collection
- `v234`: legal marital status
- `v105`: whether believes competition is good or harmful (1 to 10 scale)
- `v102`: political ideology, left to right, (1 to 10 scale)

We will treat age, the political ideology scale and the competition-attitude scale as numeric.

```{r}
### Binomial model

mod1 <- glm(v31 ~ v225 + v227 + v234 + as.numeric(age) + as.numeric(v105) + as.numeric(v102), family = "binomial", data = evs)
jtools::summ(mod1, digits =3)
```

We could also make some changes to the reference categories of some of the variables to make the interpretation of the results easier. For example, we could invert the reference category for the dependent variable so that we are modelling positive attitudes to trust rather than mistrust. We could also change the reference category for the martial status variable if we wanted to. For example: 

```{r}
evs <- within(evs, v31 <- relevel(v31, ref = "can not be too careful"))
evs <- within(evs, v234 <- relevel(v234, ref = "never married and never registered partnership"))
```

Now try to refit the previous regression (but give it a different name so we can compare!)




## Ordinal logistic regression

Let's now how much one trusts 'people they meet for the first time' (`v35`) using the same set of predictors:

```{r}
### Ordinal model; using the VGAM package

m_ord <- vglm(as.ordered(v35) ~ v225 + v227 + v234 + as.numeric(age) + as.numeric(v105) + as.numeric(v102), family = propodds, data = evs)

summary(m_ord)

```
Let's compare the results to the one's we would get using a function from another package:

```{r}
### Ordinal model; using the 'ordinal' package

m_ord2 <- clm(as.ordered(v35) ~ v225 + v227 + v234 + as.numeric(age) + as.numeric(v105) + as.numeric(v102), link="logit", data = evs)

summary(m_ord2)

```

Another fitting option exists, but it doesn't always work well:

```{r}
### Ordinal model; using the MASS package

m_ord3 <- polr(as.ordered(v35) ~ v225 + v227 + v234 + as.numeric(age) + as.numeric(v105) + as.numeric(v102), method = "logistic", data = evs)

## Unfortunately, the polr algorithm fails to find suitable starting values and cannot fit the model; we'll just leave it.

```

## Fitting the model using multinomial regression

The criteria or proportional odds is usually very demanding and the data often fails to satisfy it. In such cases, it is better to rely on a multinomial model which doesn't account for the ordering in the dependent variable.


```{r}
### Multinomial model; using the VGAM package

m_mlogit <- vglm(v35 ~ v225 + v227 + v234 + as.numeric(age) + as.numeric(v105) + as.numeric(v102), family = multinomial, data = evs)

summary(m_mlogit)
```

Reading the output is tedious. Another modelling option lets us to then use a nicer visualising function to summarize the model:

```{r}
### Multinomial model; using the 'nnet' package
m_mlogit2 <- multinom(v33 ~ v225 + v234 + v227 + as.numeric(age) + as.numeric(v102) + as.numeric(v105), data = evs)

```



```{r}
stargazer(m_mlogit2, type = "text")
```

