---
title: "Advanced social research methods 1"
subtitle: "SSC7001M"
author: "Dr. Chris Moreh<br> York St. John University"
#institute: "York St. John University"
date: "2021"
output:
  xaringan::moon_reader:
    # self_contained: true ## create self-contained slides (https://yihui.org/en/2021/06/self-contained-xaringan/)
    # seal: false ## hide the deck-title slide
    css: ["css/rutgers-rmarkdown.css", "css/rutgers-fonts_og.css"]
     # - assets/css/tachyons.min.css
     # - assets/css/animate.css
     # - assets/css/moffitt-xaringan.css
     #  - assets/css/moffitt-colors.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      # titleSlideClass: ["middle", "right"]
      # ratio: "16:9"
---
class: inverse, center, middle

```{r setup, include=FALSE} 
knitr::opts_chunk$set(
  fig.width = 10, fig.height = 6, fig.retina = 2,
  warning = FALSE, message = FALSE
)
```  

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(
  base_color = "#293352",
#  header_font_google = google_font("Josefin Sans"),
#  text_font_google   = google_font("Montserrat", "300", "300i"),
#  code_font_google   = google_font("Fira Mono")
)
# Course logo colour match
# "#C4961A", "#FC4E07", "#D16103", "#52854C", "#293352"
```

```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_progress_bar(color = "#0051BA", location = "bottom")
```
  
```{r required-packages, eval=FALSE, include=FALSE}
devtools::install_github("gadenbuie/metathis@05f8ba1f")
devtools::install_github("gadenbuie/countdown@5c895d980")
devtools::install_github("gadenbuie/xaringanExtra")
install.packages("moffitdocs")        #<< no longer available?
install.packages("xaringanthemer")  
```

# .blue[Week 1]

--

## Welcome to the module!

--

### ... and an introduction to statistical modelling for social researchers

---
class: top, left

# Session outline

--

## In the first half, I'll speak a lot:

--

### - Why quantitative sociology?
--
 .lightgrey[(have you looked at the motivational chart on Moodle and the associated reading?)]
 
--
### - Why **R** and **RStudio**?
--
 .lightgrey[(spoiler: something to do with what Freese an colleagues say in the two assigned readings?)]
 
--
### - Why statistical modelling?
--
 .lightgrey[(turning to hardcore ROS here)]

--
### - So what is this module about?
--
 .lightgrey[(Moodle page and all that)]

--
### - And how to do well on this module?
--
 .lightgrey[(looked at the Assessment Brief?)]

---
# Session outline
## In the second half, I'll relax with my .orange[coffee], read the news, check my Facebook, like some cat photos...
--
 while you:

--
#### - Fire up **RStudio**
--
 .lightgrey[and make a list of error messages that we need to report to the IT team]

--
#### - Practice some basic **R** coding
--
 .lightgrey[from the R4DS book]

--
#### - Create an **Rmarkdown** document 
--
 .lightgrey[(tip: the automatic template created is actually a mini-tutorial)]

--
#### - Locate and download the replication documentation associated with the article by Schwemmer and Wieczorek (2020)
--
 .lightgrey[ and check out the two documents titled *soc_journals_replication.Rmd* and *soc_journals_replication.nd.html*]

--
#### - You get all excited about **R** and stay here all night practicing
--
 .lightgrey[???]

---
layout: true
# Why quantitative sociology?

---
# Why R and RStudio?
--
 
---
# Why statistical modelling?
--



---
# What is this module about?
--


---
# How to do well on this module?
--


