---
title: "Advanced Research Methods 1"
subtitle: "SSC7001M"
author: "Dr. Chris Moreh<br> York St. John University"
#institute: "York St. John University"
date: "2021"
output:
  xaringan::moon_reader:
    # self_contained: true ## create self-contained slides (https://yihui.org/en/2021/06/self-contained-xaringan/)
    # seal: false ## hide the deck-title slide
    css: ["css/rutgers-rmarkdown.css", "css/rutgers-fonts_og.css"]
     # - assets/css/tachyons.min.css
     # - assets/css/animate.css
     # - assets/css/moffitt-xaringan.css
     #  - assets/css/moffitt-colors.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      # titleSlideClass: ["middle", "right"]
      # ratio: "16:9"
classoption: xcolor=table
header-includes:
      - \usepackage{booktabs}
      - \usepackage{tabu}
      - \usepackage{threeparttable}
      - \usepackage{threeparttablex}
      - \usepackage[normalem]{ulem}
---

class: inverse, center, middle

```{r setup, include=FALSE} 
knitr::opts_chunk$set(
  fig.width = 10, fig.height = 6, fig.retina = 2,
  warning = FALSE, message = FALSE, options(digits = 5)
)
```  

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(
  base_color = "#293352",
#  header_font_google = google_font("Josefin Sans"),
#  text_font_google   = google_font("Montserrat", "300", "300i"),
#  code_font_google   = google_font("Fira Mono")
)
# Course logo colour match
# "#C4961A", "#FC4E07", "#D16103", "#52854C", "#293352"
```

```{r xaringanExtra, echo = FALSE}
# devtools::install_github("gadenbuie/xaringanExtra")
xaringanExtra::use_progress_bar(color = "#0051BA", location = "bottom")
```
  
```{r required-packages, eval=FALSE, include=FALSE}
devtools::install_github("gadenbuie/metathis@05f8ba1f")
devtools::install_github("gadenbuie/countdown@5c895d980")
devtools::install_github("gadenbuie/xaringanExtra")
install.packages("moffitdocs")        #<< no longer available?
install.packages("xaringanthemer")  
```

```{r include=FALSE}
pacman::p_load(tidyverse, haven, kableExtra, knitr, ggformula, jtools, interactions)
```


# .blue[Week 6]

--

## Prediction and causality

---
# Prediction vs. causality

- **Predictive inference**: given the values of several inputs ( $X$), a fitted
model allows us to predict $y$, typically considering the $n$ data points as a simple random sample from a hypothetical infinite “superpopulation”

- **Causal inference**: what would happen to an outcome $y$ as a result of a treatment, intervention, or exposure $z$, given pre-treatment information $x$. 

- Causal effects seen as a comparison between different *potential outcomes* of what might have occurred under different scenarios (counterfactual states) (see also Morgan and Winship (2014))

- We assume the existence of potential outcome random variables ( $Y^{1 (treated)}$, and $Y^{0 (control)}$) that are defined over all individuals in the population of interest

--

## $$ \delta_i = y_i^1 - y_i^0 $$

** $\delta$ is read "delta" and usually denotes change in statistical equations

---
# The problem of causality

- The fundamental problem of causal inference: we can never observe both potential outcomes $y_i^1$ and  $y_i^0$ for the same  individual. Therefore, we cannot determine any individual causal effect without further assumptions.

- Assuming treatments are randomly assigned, we can estimate an *average* causal effect, but this still only applies to individuals if we are willing to also assume that effects are constant across people

- Assessing causal mechanisms require specific data types that are able to capture causality.

---
# Causal data

- **Experimental data**: collected from an experimental research design, in which a treatment variable, or a causal variable of interest, is manipulated in order to examine its causal effects on an outcome variable

   - **Randomized experiments**: a.k.a. 'Randomized Controlled Trials' (RCTs); in which researchers randomly assign the receipt of treatment. Regarded as the gold standard for establishing causality because it enables researchers to isolate the effects of a treatment variable and quantify uncertainty
   
   - **Natural experiments**: “a real world situation that produces haphazard assignment to a treatment” (Rosenbaum 2010: 67)

Examples:

- Bertrand, M. and Mullainathan, S. (2004) “Are Emily and Greg more employable than Lakisha and Jamal? A field experiment on labor market discrimination.” American Economic Review, vol. 94, no. 4, pp. 991–1013.
- Ladd, J.M. and Lenz, G.S. (2009) ‘Exploiting a Rare Communication Shift to Document the Persuasive Power of the News Media’, American Journal of Political Science, 53(2), pp. 394–410.

---
# Causal data

- **Observational data**: naturally occurring events. Internal validity likely compromised because of possible selection bias. But findings more generalizable because we can examine the treatments that are implemented among a relevant population in a real-world environment.

  - **Longitudinal**: Observing units (individuals, countries etc.) several times; longer time-series for aggregate units (e.g. country-level measures across time). In such designs, causal effects can be approximated by assessing change over time within units
  
  - **Cross-sectional**: Data collected only at one time-point for each unit (most of the data we have explored so far)

Examples:

- Österman, M. (2021) ‘Can We Trust Education for Fostering Trust? Quasi-experimental Evidence on the Effect of Education and Tracking on Social Trust’, Social Indicators Research, 154(1), pp. 211–233.
- Bauer, P.C. (2015) ‘Negative Experiences and Trust: A Causal Analysis of the Effects of Victimization on Generalized Trust’, European Sociological Review, 31(4), pp. 397–417.

---
# Example 1: Racial discrimination in the labour-market


## Based on: 
### Bertrand, M. and Mullainathan, S. (2004) “Are Emily and Greg more employable than Lakisha and Jamal? A field experiment on labor market discrimination.” American Economic Review, vol. 94, no. 4, pp. 991–1013.

## Import data:
```{r message=FALSE, warning=FALSE}
resume <- read_csv("https://cgmoreh.github.io/SSC7001M/data/resume.csv")
```

---
```{r paged.print=TRUE}
resume 
```

```{r}
summary(resume)
```

---
## Call-back by racial group
```{r}
race_call_tab <-
  resume %>%
  group_by(race, call) %>%
  count()
kable(race_call_tab)
```

---
## Call-back rates by race

```{r}
race_call_rate <-
  race_call_tab %>%
  group_by(race) %>%
  mutate(call_rate =  n / sum(n)) %>%
  filter(call == 1) %>%
  select(race, call_rate)
kable(race_call_rate)
```

---
## Call-back rates by gender and race
```{r}
resume_race_sex <-
  resume %>%
  group_by(race, sex) %>%
  summarise(call = mean(call))
kable(resume_race_sex)
```

```{r}
resume_sex <-
  resume_race_sex %>%
  ungroup() %>%
  spread(race, call)
kable(resume_sex %>% mutate(call_diff = white - black))
```
---
## Using regression

A race only model:
```{r}
# Build the model
race_reg <- lm(call ~ race, data=resume)

# Describe the model
jtools::summ(race_reg, digits=3)
```

---
## Using regression
A race and gender model:
```{r}
# Build the model
race_reg <- lm(call ~ race + sex, data=resume)

# Describe the model
jtools::summ(race_reg, digits=3)
```
---
## Using regression
A race and gender model with interaction:
```{r}
# Build the model
race_reg <- lm(call ~ race*sex, data=resume)

# Describe the model
jtools::summ(race_reg, digits=3)
```

---
# Exercises

## 2. Can the media persuade people to vote differently? 

## 3. Osterman data exercise

See and follow the .Rmd file